---
title: "Comparison of isomiR importance"
author: "Lorena Pantano"
date: "`r Sys.Date()`"
output:
    md_document:
        variant: markdown_github
        toc: false
params:
    data: "data/data_gff.rda"
    outputDir: "."
---

```{r aes, results='hide'}
library(knitr)
library(ggplot2)
library(pheatmap)

# Set seed for reproducibility
set.seed(1454944673L)

opts_chunk[["set"]](
    autodep = TRUE,
    bootstrap.show.code = FALSE,
    cache = TRUE,
    cache.lazy = TRUE,
    dev = c("png", "pdf"),
    error = TRUE,
    fig.height = 6,
    fig.retina = 2L,
    fig.width = 8,
    highlight = TRUE,
    message = FALSE,
    prompt = TRUE,
    # formatR required for tidy code
    tidy = TRUE,
    warning = FALSE,
    results = 'asis')

theme_set(
    theme_light(base_size = 11L))
theme_update(
    legend.justification = "center",
    legend.position = "bottom")
```

# Data analyzed

* equimolar Tewari: bcbio, miRge and razer3
* custom equimolar Tewari: razer3
* custom Van Dijk: razer3
* plasma: bcbio,  miRge

## Filtering applied:

* Only using sequences that map to miRNAs that are in the miRXplore sample or spike-ins defined by the project 
* miRNAs that didn't contain one sequence being inside the miRXplore or spike-ins space  were removed
* Only sequences that mapped once to a reference were kept

# Conclusion

The majority of miRNAs had the reference sequences as top expressed.

The majority of isomiRs correspon to trimming events.

The majority are losses/truncation.

4N generetes more trimming events.

# Results

```{r, cahce = FALSE}
library(tidyverse)
library(ggplot2)

load(params$data)
```

* From the miRNAs where the reference is the most expressed, what is the percentage of the isomiR expression from the total miRNA expression

```{r}
equimolar %>% 
    filter(ref_is_1 == 1) %>%
    ggplot(aes(x = short, fill = pct_cat)) +
    geom_bar() +
    facet_wrap(~iso, nrow = 4, scales = "free_y") +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
    ggtitle("equimolar - bcbio")

equimolar_mirge %>% 
    filter(ref_is_1 == 1) %>%
    group_by(short, mi_rna) %>% 
    ggplot(aes(x = short, fill = pct_cat)) +
    geom_bar() +
    facet_wrap(~iso, nrow = 4, scales = "free_y") +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
    ggtitle("equimolar - mirge2.0")

equimolar_razer3 %>% 
    filter(ref_is_1 == 1) %>%
    group_by(short, mi_rna) %>% 
    ggplot(aes(x = short, fill = pct_cat)) +
    geom_bar() +
    facet_wrap(~iso, nrow = 4, scales = "free_y") +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
    ggtitle("equimolar - razer3")

custom %>% 
    filter(ref_is_1 == 1) %>% 
    ggplot(aes(x = short, fill = pct_cat)) +
    geom_bar() +
    facet_wrap(~iso, nrow = 4, scales = "free_y") +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
    ggtitle("custom - razer3")

vandijk %>% 
    filter(ref_is_1 == 1) %>% 
    ggplot(aes(x = short, fill = pct_cat)) +
    geom_bar() +
    facet_wrap(~iso, nrow = 4, scales = "free_y") +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
    ggtitle("VanDijk - razer3")


plasma %>% 
    filter(ref_is_1 == 1) %>%
    ggplot(aes(x = short, fill = pct_cat)) +
    geom_bar() +
    facet_wrap(~iso, nrow = 4, scales = "free_y") +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
    ggtitle("plasma - bcbio")

```

Same thing but using seqThreshold to keep sequences.

```{r}
equimolar %>% 
    filter(ref_is_1 == 1, value > threshold) %>%
    ggplot(aes(x = short, fill = pct_cat)) +
    geom_bar() +
    facet_wrap(~iso, nrow = 4, scales = "free_y") +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
    ggtitle("equimolar - bcbio")

equimolar_mirge %>% 
    filter(ref_is_1 == 1, value > threshold) %>%
    group_by(short, mi_rna) %>% 
    ggplot(aes(x = short, fill = pct_cat)) +
    geom_bar() +
    facet_wrap(~iso, nrow = 4, scales = "free_y") +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
    ggtitle("equimolar - mirge2.0")

equimolar_razer3 %>% 
    filter(ref_is_1 == 1, value > threshold) %>%
    group_by(short, mi_rna) %>% 
    ggplot(aes(x = short, fill = pct_cat)) +
    geom_bar() +
    facet_wrap(~iso, nrow = 4, scales = "free_y") +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
    ggtitle("equimolar - razer3")

custom %>% 
    filter(ref_is_1 == 1, value > threshold) %>% 
    ggplot(aes(x = short, fill = pct_cat)) +
    geom_bar() +
    facet_wrap(~iso, nrow = 4, scales = "free_y") +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
    ggtitle("custom - razer3")

vandijk %>% 
    filter(ref_is_1 == 1, value > threshold) %>% 
    ggplot(aes(x = short, fill = pct_cat)) +
    geom_bar() +
    facet_wrap(~iso, nrow = 4, scales = "free_y") +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
    ggtitle("VanDijk - razer3")


plasma %>% 
    filter(ref_is_1 == 1, value > threshold) %>%
    ggplot(aes(x = short, fill = pct_cat)) +
    geom_bar() +
    facet_wrap(~iso, nrow = 4, scales = "free_y") +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
    ggtitle("plasma - bcbio")

```


Same thing but normalizing by number of total sequences detected in each sample.

```{r}
prepare =  . %>% filter(ref_is_1 == 1) %>%
    dplyr::count(short, pct_cat, iso) %>% 
    group_by(short) %>% 
    mutate(pct_total = n/sum(n)*100)

equimolar %>% 
    prepare() %>% 
    ggplot(aes(x = short, fill = pct_cat, y = pct_total)) +
    geom_bar(stat = "identity") +
    facet_wrap(~iso, nrow = 4, scales = "free_y") +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
    ggtitle("equimolar - bcbio")

equimolar_mirge %>% 
    prepare() %>% 
    ggplot(aes(x = short, fill = pct_cat, y = pct_total)) +
    geom_bar(stat = "identity") +
    facet_wrap(~iso, nrow = 4, scales = "free_y") +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
    ggtitle("equimolar - mirge2.0")

equimolar_razer3 %>% 
    prepare() %>% 
    ggplot(aes(x = short, fill = pct_cat, y = pct_total)) +
    geom_bar(stat = "identity") +
    facet_wrap(~iso, nrow = 4, scales = "free_y") +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
    ggtitle("equimolar - razer3")

custom %>% 
    prepare() %>% 
    ggplot(aes(x = short, fill = pct_cat, y = pct_total)) +
    geom_bar(stat = "identity") +
    facet_wrap(~iso, nrow = 4, scales = "free_y") +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
    ggtitle("custom - razer3")

vandijk %>% 
    prepare() %>% 
    ggplot(aes(x = short, fill = pct_cat, y = pct_total)) +
    geom_bar(stat = "identity") +
    facet_wrap(~iso, nrow = 4, scales = "free_y") +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
    ggtitle("VanDijk - razer3")

plasma %>% 
    prepare() %>% 
    ggplot(aes(x = short, fill = pct_cat, y = pct_total)) +
    geom_bar(stat = "identity") +
    facet_wrap(~iso, nrow = 4, scales = "free_y") +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
    ggtitle("plasma - bcbio")

```

* From the miRNAs where the reference is the most expressed, what is the percentage of the isomiR expression from the total miRNA expression, only looking at iso_5p and iso_3p

```{r}
equimolar %>% 
    filter(ref_is_1 == 1, iso_shift_nt!="0", (iso=="&shift5p&" | iso=="&shift3p")) %>%
    ggplot(aes(x = short, fill = pct_cat)) +
    geom_bar() +
    facet_wrap(~iso_shift_nt, nrow = 4, scales = "free_y") +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
    ggtitle("equimolar - bcbio")

plasma %>% 
    filter(ref_is_1 == 1, iso_shift_nt!="0", (iso=="&shift5p&" | iso=="&shift3p")) %>%
    ggplot(aes(x = short, fill = pct_cat)) +
    geom_bar() +
    facet_wrap(~iso_shift_nt, nrow = 4, scales = "free_y") +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
    ggtitle("plasma - bcbio")

custom %>% 
    filter(ref_is_1 == 1, iso_shift_nt!="0", (iso=="&shift5p&" | iso=="&shift3p")) %>%
    ggplot(aes(x = short, fill = pct_cat)) +
    geom_bar() +
    facet_wrap(~iso_shift_nt, nrow = 4, scales = "free_y") +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
    ggtitle("custom - razer3")

vandijk %>% 
    filter(ref_is_1 == 1, iso_shift_nt!="0", (iso=="&shift5p&" | iso=="&shift3p")) %>%
    ggplot(aes(x = short, fill = pct_cat)) +
    geom_bar() +
    facet_wrap(~iso_shift_nt, nrow = 4, scales = "free_y") +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
    ggtitle("VanDijk - razer3")

equimolar_mirge %>% 
    filter(ref_is_1 == 1, iso_shift_nt!="0", (iso=="&shift5p&" | iso=="&shift3p")) %>%
    ggplot(aes(x = short, fill = pct_cat)) +
    geom_bar() +
    facet_wrap(~iso_shift_nt, nrow = 4, scales = "free_y") +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
    ggtitle("equimolar - mirge2.0")

```

